Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-28 trunk/paper/drinking.tex

\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases \cite{}.

@@

Here, we present the foundations for an ontology of physiological
phenotypes.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005a}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for the {\tt MP:0009007} term from
the Mammalian Phenotype) it could be used to provide an equivalence
statement between that class and the above PATO-based description.

@@

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
  \caption{Overview over ...\label{fig:patterns}}
\end{figure}

@@

\subsection{Pre- vs. post-composed ontology}
Pre-composed or software support for pattern?

\subsection{Formalization}
Formal, axiomatic system for process qualities, maybe based on PSL,
FOIS paper 2010

one example:...

\subsection{Conclusion}

>>>>

\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases
\cite{Schofield2011}.

@@

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists...

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for the {\tt MP:0009007} term from
the Mammalian Phenotype) it could be used to provide an equivalence
statement between that class and the above PATO-based description.

@@

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}

  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

@@

\subsection{Pre- vs. post-composed ontology}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify use
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result is a large ontology in which classes for phenotypes
are {\em pre-composed}: they are named and defined within an OWL
ontology. However, the large size of the resulting ontology may impair
its utility for data annotation and integration, and software tools
may not always support such very large ontologies. The alternative to
pre-composing all possible phenotype classes using the patterns we
describe is to dynamically generate appropriately defined classes at
the time at which they are being used. To achieve this goal, software
must be developed to support ontology users in applying these patterns
and generate the appropriate class description when required.

% \subsection{Future work}
% Formal, axiomatic system for process qualities, maybe based on PSL,
% FOIS paper 2010

% one example:...

% \subsection{Conclusion}
